Kepler's Demographic Cliff: Ages of Young Stars and Planets in the Kepler Field

### 2024-05-06 paper/ms.tex
update ms

\subsubsection{Rotation vs.~Lithium Ages}

@@

In the entire sample (i.e.~without imposing an quality cuts)
this yielded \allagesyesconsistent\ consistent
cases, \allagesmaybeconsistent\ ``maybe consistent'' cases, and
\allagesnoconsistent\ inconsistent.  Restricting our attention to the
$\lesssim$1\,Gyr stars, for which lithium ages are more constraining,
the respective counts are \minageltonegyryesconsistent,
\minageltonegyrmaybeconsistent, and \minageltonegyrnoconsistent.
Rephrased, in the overall sample, \fracconsistentallages\% of stars
have consistent rotation and lithium-based ages
(\fracpotentiallyconsistentallages\% potentially consistent).  In the
sample of sub-gigayear stars, \fracconsistentminageltonegyr\% of the
ages are consistent (\fracpotentiallyconsistentminageltonegyr\%
potentially consistent).  We feel it necessary to distinguish
``maybe'' consistent cases because of lingering uncertainties in how
the dispersion of the lithium data evolve over the first gigayear.

In
Appendix~\ref{app:inconsistent}, we discuss the
systems for which at least one age indicator
reports a sub-gigayear age in detail.
All but one are simply cases for which the star
is either evolved or else an unresolved binary.
The one exception is Kepler-786 (KOI-1199), an early K dwarf with 
$t_{\rm Li}$=\kepseveneightsix\,Myr, which would predict an obvious rotation signal.
No photometric signal is detected; other age indicators in the
spectrum similarly suggest that the star may not in fact be young.

{\bf FIXME BRIEF SUMMARY, e.g. N of them are because lithium is garbage,
	M are because rotation is weird}.

\subsubsection{New Young Systems}
%
Whenever both rotation and lithium find a star to be in a
narrowly-constrained age range, confidence in the true age greatly
increases, due to the independence of the two measurement techniques.
The empirical precision for the inferred ages as a function of $P_{\rm
rot}$, $EW_{\rm Li}$, and $T_{\rm eff}$ has been mapped out by
\citet{Bouma_2023} and \citet{Jeffries_2023} for rotation and lithium
respectively.  The region of greatest overlap is dictated by the
timescale for lithium depletion: for stars with temperatures between
4500-5500\,K with ages $\lesssim$500\,Myr, we can hope to obtain
independent two-sided constraints.

The youth of most of these systems, to our knowledge, is being
well-constrained for the first time in this work.  Here we limit
discussion to systems with $t_{\rm gyro}$$<$500\,Myr not in the clusters
mentioned above.

@@

Of these planets, one specific highlight is Kepler-1933b, a 4.9\,day
1.0\,$R_\oplus$ Earth-sized planet.  Its rotation-based age of
$304^{+96}_{-112}$\,Myr agrees within $1\sigma$ of its lithium age,
though at half the relative uncertainty.  Based on these lines of
evidence, this system may be the youngest Earth-sized planet known,
comparable to systems such as HD~63433d \citep[1.1\,$R_\oplus$,
$414\pm23$\,Myr][]{2024AJ....167...54C}.

>>>>

\subsubsection{Comparing Rotation and Lithium Ages}

@@

In the entire sample (i.e.~without imposing any quality cuts) this
yielded \allagesyesconsistent\ consistent cases,
\allagesmaybeconsistent\ ``maybe consistent'' cases, and
\allagesnoconsistent\ inconsistent cases.  Rephrased, in the overall
sample, \fracconsistentallages\% of stars have consistent rotation and
lithium-based ages (\fracpotentiallyconsistentallages\% potentially
consistent).

Appendix~\ref{app:inconsistent} describes all cases of ``discrepant''
confirmed planets for which a sub-gigayear age is reported by at least
one age indicator.  All but two systems are
either evolved stars or else unresolved binaries, and are
automatically flagged as such.  The two exceptions are Kepler-786
(KOI-1199), and Kepler-1 (TrES-2).  Kepler-786 is an early K dwarf with
$t_{\rm Li}$=\kepseveneightsix\,Myr, which would imply that
rotation should be detectable.  However no photometric rotation signal is detected; other age
indicators in the spectrum similarly suggest that the star may not in
fact be young.  TrES-2 is therefore a ``lithium-young'' system for
which we believe the high lithium EW may be
a statistical fluke.

%Restricting our attention to the
%$\lesssim$1\,Gyr stars, for which lithium ages are more constraining,
%the respective counts are \minageltonegyryesconsistent,
%\minageltonegyrmaybeconsistent, and \minageltonegyrnoconsistent.
%Rephrased, in the overall sample, \fracconsistentallages\% of stars
%have consistent rotation and lithium-based ages
%(\fracpotentiallyconsistentallages\% potentially consistent).  In the
%sample of sub-gigayear stars, \fracconsistentminageltonegyr\% of the
%ages are consistent (\fracpotentiallyconsistentminageltonegyr\%
%potentially consistent).  We feel it necessary to distinguish
%``maybe'' consistent cases because of lingering uncertainties in how
%the dispersion of the lithium data evolve over the first gigayear.

\subsubsection{New Young Systems}
%
Whenever both rotation and lithium find a star to be in a
narrowly-constrained age range, confidence in the true age
increases, because the two techniques are independent.
The region of greatest overlap between rotation and lithium ages in
the spaces of $P_{\rm rot}$, $EW_{\rm Li}$, and $T_{\rm eff}$ is
dictated by the timescale for lithium depletion: for stars with
temperatures between 4500-5500\,K with ages $\lesssim$500\,Myr, we can
hope to obtain independent two-sided constraints
\citep{Bouma_2023,Jeffries_2023}.

The following systems, to our knowledge, are being
well-constrained for the first time in this work.  Here we limit
discussion to systems with $t_{\rm gyro}$$<$500\,Myr not in the clusters
mentioned above.
{\bf FIXME TODO: this section is boring.  come up with a better way to do this,
and maybe automate}

@@

Of these planets, one specific highlight is Kepler-1933b, a 4.9\,day
1.0\,$R_\oplus$ Earth-sized planet.  Its rotation-based age of
$304^{+96}_{-112}$\,Myr agrees within $1\sigma$ of its lithium age,
though at half the relative uncertainty.  This system may be the
youngest Earth-sized planet currently known, comparable to systems
such as HD~63433d \citep[1.1\,$R_\oplus$,
$414\pm23$\,Myr][]{2024AJ....167...54C}.
